Measurement and Probability in Fuzzy Quantum Theories

### 2016-11-27 proposal/prop.tex
prop

The nature of quantum mechanics realized in terms of discrete fields is
amenable to direct computational exploration for reasonable field sizes. We
have established a set of utility libraries in Mathematica that supports a
wide variety of such numerical experiments, with Fig.~\ref{fig_qm} being an
example of the typical output.

Concurrently to this effort, we are also building on our previous
work~\cite{qhaskell,DBLP:journals/entcs/AltenkirchGVS07,DBLP:journals/entcs/Vi\
zzottoCS08,DBLP:conf/wollic/VizzottoBS09,DBLP:conf/sbmf/VizzottoLS09,1166040}
to build semantically well-founded implementations of the computational
models investigated in this proposal.  The aim of these implementations is
two-fold: (i) to provide an experimental testbed that complements the
Mathematica libraries in which discrete quantum algorithms can be simulated;
and (ii) to establish connections between discrete quantum programming
concepts and classical programming concepts.
 We would also like to simulate physical phenomena by using DQCq
 \cite{OrtizEtAl2001,SommEtAl2002}.

\paragraph*{Proposed Activities.}
\begin{itemize}
\item We plan to make our software available for educational and
  research purposes.
\end{itemize}

@@

\noindent (A) {\em Integration of research and education:} Courses on
advanced computational models and quantum information will be further
developed at Indiana University to communicate the basic concepts and tools
discussed in this proposal. Dr. Ortiz taught courses on ``Quantum information   
and Computation'' and Dr. Sabry has taught courses on ``Quantum Computing''
for graduate students.  These courses were highly appreciated by the students
as noted by their evaluations in making the material extremely accessible and
interesting.  Dr.\ Hanson has taught courses on scientific visualization and
mathematical modeling with emphasis on high-dimensional mathematics.  His
numerous Siggraph course presentations are embodied in his 2006 book {\it       
Visualizing Quaternions\/}, which remains the definitive treatment of the
subject for computer science and data analysis applications and has sold
thousands of copies. As a part of Dr.\ Hanson's commitment to exposing and
explaining the elements of higher mathematics to the public, he has recently
led the design of a free iPhone App, {\it 4Dice\/}, that for the first time
makes available an interactive treatment of a correctly imaged simulation of
the appearance of a hypercube, realizing interactively his YouTube animation
of the same name.  This framework is now being generalized for a broad family
of iPhone Apps to permit the hands-on study of high-dimensional mathematics.
The PIs plan to engage high school students or undergraduates in the
implementation of some of the ideas and the exploration of finite geometry.
The PIs have a long standing commitment to K-12 science and math education
(see also (C) below). The collaborative group efforts will provide students,
as well as the PIs with a unique learning experience. The broadened
perspectives gained through those efforts will contribute to the PIs
effectiveness as teachers and mentors. They also will enable to better
prepare students and postdocs for productive careers in an increasingly
competitive global economy. Students and postdocs are encouraged to pursue
collaborative research leading to a broader scientific program. This is
highlighted by the PIs students' close contact with LANL, the Perimeter
Institute, and the Institute for Quantum Computing at Waterloo.

@@

\medskip
\noindent (C) {\em Enhancing infrastructure and outreach:} Dr. Ortiz has
pursued broad outreach by giving talks on astronomy/astrophysics in the local
schools at Bloomington, Indiana, and by an interview in the public radio on
quantum information. Dr. Ortiz is also guiding a high school student in work
on interferometry in two and three slit experiments. Dr. Ortiz also guides
undergraduate research. Dr.\ Hanson plans to add interactive exploration of
the geometry of quantum computing (see Fig.~\ref{fig_qm}) to the suite of
educational iPhone Apps currently being developed. All source code produced
by this project will be made freely available via the web, and the results
will be disseminated via technical reports, conference papers, and journal
articles.  The software and concepts we have developed in the past have been
employed by a wide variety of commercial and noncommercial users.  Thus, we
expect that the successful completion of the proposed research will
contribute to the understanding of quantum and semi quantum computation. The
educational benefits that will flow from this proposal will extend beyond the
students of the participating research groups. The knowledge gained through
participation in the proposed research will further strengthen the
effectiveness of these outreach activities.

@@

There was no recent NSF support for Dr. Ortiz and Dr. Hanson. In the
summer of 2011, Dr. Sabry received in collaboration with two of his
colleagues two NSF grants: \emph{Effectful Software Contracts} with
Dr. Amal Ahmed and \emph{Reversible Concurrency} with Dr. Geoffrey
Brown. Both proposals are in their initial stages and no significant
results have appeared yet.

Over the past several years, the PI, together with other IU faculty
members and students, has formed a group whose main interest is the
study of paradigms of computing that either arise in nature or are
inspired by natural phenomena. The PI is especially interested in how
the laws of physics can inspire and guide our abstract computational
models, as well as exposing new insights into the computational
framework behind the natural world and the computability of physical
theory. Our group has been active in several areas related to the
theory behind quantum computation. Particular areas include:
mathematical foundations of quantum computation and
complexity~\cite{AGVS2007,VAS2006},
adiabatic quantum computing, quantum logic, classical and quantum
reversible programming languages~\cite{Sabry:2003:MQC:871895.871900},
analog computation and its limits, and optical computing. During the
last year, our discussions have revolved around formulating variants
of quantum mechanics that are computable and that are as close to
possible to classical computing. Several preliminary results have
emerged from our collaboration~\cite{usat,finiteQC,geometry2013,DQT2014}, and the group
is very excited about the potential implications of those
results. This proposal brings forth a particular aspect of this
programme related to its applications to problem areas that have
traditionally been outside the conceptual framework of programming
language research.

In the past five years, Prof. Sabry received one NSF grant (award
number 1217454) titled \emph{Information Effects} for the amount of
\$275,319.00 with start and end dates of 1 September 2012 -- 31 August
2016.

>>>>

The nature of quantum mechanics realized in terms of discrete fields
and finitely-communicable probability estimates is amenable to direct
computational exploration for reasonable field sizes and observational
accuracy. We have established a set of utility libraries in
Mathematica that supports a wide variety of such numerical
experiments. Concurrently to this effort, we are also building on our
previous
work~\cite{VDS2009,AGVS2007,AGVS2005,VAS2006,finiteQC,Sabry:2003:MQC:871895.871900}
to build semantically well-founded implementations of the
computational models investigated in this proposal.  The aim of these
implementations is two-fold: (i) to provide an experimental testbed to
simulate physical phenomena (e.g.~\cite{OrtizEtAl2001,SommEtAl2002})
that complements the Mathematica libraries in which quantum algorithms
can be simulated; and (ii) to establish connections between quantum
programming concepts and classical programming concepts.  We plan to
make our software available for educational and research purposes.

@@

\noindent (A) {\em Integration of research and education:} Courses on
advanced computational models and quantum information will be further
developed at Indiana University to communicate the basic concepts and
tools discussed in this proposal. Dr. Ortiz taught courses on
``Quantum information and Computation'' and Dr. Sabry has taught
courses on ``Quantum Computing'' for graduate students.  These courses
were highly appreciated by the students as noted by their evaluations
in making the material extremely accessible and interesting.  The PIs
plan to engage high school students or undergraduates in the
implementation of some of the ideas and the exploration of finite
geometry.  The PIs have a long standing commitment to K-12 science and
math education (see also (C) below). The collaborative group efforts
will provide students, as well as the PIs with a unique learning
experience. The broadened perspectives gained through those efforts
will contribute to the PIs effectiveness as teachers and mentors. They
also will enable to better prepare students and postdocs for
productive careers in an increasingly competitive global
economy. Students and postdocs are encouraged to pursue collaborative
research leading to a broader scientific program. This is highlighted
by the PIs students' close contact with LANL, the Perimeter Institute,
and the Institute for Quantum Computing at Waterloo.

@@

\medskip
\noindent (C) {\em Enhancing infrastructure and outreach:} Dr. Ortiz
has pursued broad outreach by giving talks on astronomy/astrophysics
in the local schools at Bloomington, Indiana, and by an interview in
the public radio on quantum information. Dr. Ortiz is also guiding a
high school student in work on interferometry in two and three slit
experiments. Dr. Ortiz also guides undergraduate research. All source
code produced by this project will be made freely available via the
web, and the results will be disseminated via technical reports,
conference papers, and journal articles.  The software and concepts we
have developed in the past have been employed by a wide variety of
commercial and noncommercial users.  Thus, we expect that the
successful completion of the proposed research will contribute to the
understanding of quantum and semi quantum computation. The educational
benefits that will flow from this proposal will extend beyond the
students of the participating research groups. The knowledge gained
through participation in the proposed research will further strengthen
the effectiveness of these outreach activities.

Over the past several years, the PIs, together with other IU faculty
members and students, has formed a group whose main interest is the
study of paradigms of computing that either arise in nature or are
inspired by natural phenomena. The PI is especially interested in how
the laws of physics can inspire and guide our abstract computational
models, as well as exposing new insights into the computational
framework behind the natural world and the computability of physical
theory. Our group has been active in several areas related to the
theory behind quantum computation. Particular areas include:
mathematical foundations of quantum computation and
complexity~\cite{AGVS2007,VAS2006}, adiabatic quantum computing,
quantum logic, classical and quantum reversible programming
languages~\cite{Sabry:2003:MQC:871895.871900}, analog computation and
its limits, and optical computing. During the last year, our
discussions have revolved around formulating variants of quantum
mechanics that are computable and that are as close to possible to
classical computing. Several preliminary results have emerged from our
collaboration~\cite{usat,finiteQC,geometry2013,DQT2014}, and the group
is very excited about the potential implications of those
results. This proposal brings forth a particular aspect of this
programme related to its applications to problem areas that have
traditionally been outside the conceptual framework of programming
language research.

@@

There was no recent NSF support for Dr. Ortiz. In the past five years,
Prof. Sabry received one NSF grant (award number 1217454) titled
\emph{Information Effects} for the amount of \$275,319.00 with start
and end dates of 1 September 2012 -- 31 August 2016.
